---
title: "Data manipulation"
author: "FESTUS DAVID L."
date: "2026-03-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#tidyverse in R 
#is the most popular meta-package for modern data science in R. It bundles a set of #consistent, opinionated packages that work together using the same grammar and the pipe #operator (%>% or the newer native |>).

#What you get when you load tidyverse

```{r}
install.packages("tidyverse")   # run once
library(tidyverse)
```

#This automatically loads the core eight packages (plus a few helpers):

#Package      Purpose                 Most-used functions
#dplyr        Data manipulation       filter(), select(), mutate(), arrange(),                                                summarise(), group_by()
#tidyr        Reshaping data          pivot_longer(), pivot_wider(), separate(),                                              unnest()
#ggplot2      Grammar of graphics     ggplot() + geoms (geom_point(), geom_bar(), etc.)
#readr        Fast data import        read_csv(), read_tsv(), read_excel() (via readxl)
#tibble       Modern data frames      tibble(), as_tibble(), nicer printing
#purrr        Functional programming  map(), map_dfr(), pmap()
#stringr      String manipulations    tr_detect(), str_replace(), str_glue()
#forcats      Working with factors    fct_reorder(), fct_lump(), fct_relevel()


#The “tidy” philosophy

#Tidy data = each variable is a column, each observation is a row, each value is a cell.
Everything is designed around this format → no more fighting with base R’s weird syntax.
```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

