---
title: "reading_and_writing_data"
author: "FESTUS DAVID L."
date: "2026-03-25"
output: pdf_document
---

#Reading and Writing Data in R
# Learning Objectives
#By the end of this lesson, 
#- Understand common data file formats used in data analysis
#- Read data into R from CSV, TSV, Excel, and RDS files
#- Write data from R back to disk in different formats
#- Perform quick checks to verify successful import

# Introduction: Why Data I/O Matters
#Before any analysis, you must import your data into R. After cleaning or analyzing it, #you often need to export results for sharing, reporting, or further processing.
#R provides simple, powerful functions for both tasks. This tutorial walks through the #most common ones.


#Reading Data into R
#1 Reading CSV Files
#CSV (Comma‑Separated Values) is the most common format for data exchange.

#Explanation
#- header = TRUE tells R the first row contains column names
#- stringsAsFactors = FALSE prevents automatic conversion of text into factors

```{r}
data_csv <- read.csv("data_file.csv", 
                     header = TRUE, 
                     stringsAsFactors = FALSE)
```

#Reading TSV Files
#TSV files use tabs instead of commas.
# Function: read.delim()
```{r}
data_tsv <- read.delim("data_file.tsv", header = TRUE)
```

#Reading Excel Files
#Excel files require an external package
#Explanation
#- sheet = 1 loads the first sheet
#- No need for Java or heavy dependencies

```{r}
library(readxl)
data_excel <- read_excel("data_file.xlsx", sheet = 1)
```

#Reading RDS Files
#RDS files store R objects in a compact format
#Explanation
#• 	Preserves object structure (lists, models, data frames)
#• 	Ideal for saving intermediate results
```{r}
data_rds <- readRDS("data_file.rds")
```

#Fast Reading with data.table::fread()
#For large datasets, fread() is extremely fast.

```{r}
library(data.table)
data_fast <- fread("large_file.csv")
```

# Writing Data from R to Disk
# Writing CSV Files
# Explanation
#- row.names = FALSE avoids adding an unnecessary first colum

```{r}
write.csv(data_csv, "output_file.csv", row.names = FALSE)
```

#wiring TSV files
```{r}
write.table(data_tsv, "output_file.tsv", 
            sep = "\t", 
            row.names = FALSE, 
            quote = FALSE)
```

#wiritng excel files
```{r}
library(writexl)
write_xlsx(data_excel, "output_file.xlsx")
```

#Writing RDS Files
```{r}
saveRDS(data_rds, "output_file.rds")
```

#Fast Writing with fwrite()
```{r}
fwrite(data_fast, "output_fast.csv")
```

